opXRD: Open Experimental Powder X-ray Diffraction Database

### 2025-04-28 chapters/04_our_dataset.tex
Described new xrd/structure figures

\begin{table}[!htb]
\centering
\caption{\footnotesize Overview of the contributions to the opXRD database: The availability of the chemical composition, spacegroups, lattice parameters, and atomic coordinates of the underlying samples are indicated by the columns ``Comp.'', ``Spg.'', ``Lattice'' and ``\replaced{Basis}{Atom coords}.'' respectively.}
\label{tab:merged}
\scalebox{0.82}{
\begin{adjustbox}{center}
\begin{NiceTabular}{@{}llllllll@{}}[columns-width=1.2cm]
%\begin{tabular}{@{}llccccl@{}}
\toprule
\textbf{Institution} & \textbf{No. patterns} & \textbf{Comp.} & \textbf{Spg.} & \textbf{Lattice} & \textbf{Basis} & \textbf{Research Project} \\
\midrule
CNRS        &  1052         &  \cmark   & \pchk{85}  & \cmark    &   \pchk{85}   & Diffraction data extracted from the COD \\ 
USC         &  338          &  \cmark   & \cmark     & \pchk{90} &   \xmark   & Study of CuNi and CuAl alloys          \\
HKUST(GZ)   &  520          &  \pchk{4} & \pchk{4}   & \pchk{4}  &   \pchk{4}   & Phase identification dataset  \\ 
EMPA        &  770          & \cmark    & \pchk{63}  & \xmark    &   \xmark   & Metal halide perovskites, Zn-V-N libraries \\ 
INT         &  19,796       & \xmark    & \xmark     & \xmark    &   \xmark   & Compilation of various projects        \\ 
IKFT        &  64           & \xmark    & \xmark     & \xmark    &   \xmark   & Commercial catalysts, metals, metal oxides   \\ 
LBNL        &  70,012       & \xmark    & \xmark     & \xmark    &   \xmark   & Perovskites precursors, Mn-Sb-O system  \\ 
\bottomrule
\end{NiceTabular}
\end{adjustbox}}
\end{table}

The variance of the data was analyzed using principal component analysis (PCA). PCA can be applied to datasets $X \subset \mathbb{R}^N$ to reduce the number of components needed to describe points $p \in X$ up to some tolerance in lost accuracy. In the context of PCA, the cumulative explained variance ratio is a measure of how much of the variance in the dataset $X$ can be explained using a given number of components. For a rigorous definition of PCA and the explained variance ratio, we refer to the literature \cite{Jolliffe2016}. Here, PCA was performed on datasets of X-ray diffraction patterns. These datasets $X$ are subsets of $\mathbb{R}^{N}$ with $N=512$ since each pattern $p \in X$ was standardized to have 512 intensity values spread out evenly from $0\degree$ to $180\degree$ using zero padding and interpolation with cubic splines. Hence the maximal components that could be needed to describe a dataset of diffraction data in this context is $N=512$. However, the maximal number of components is even lower for datasets that contain less than 512 patterns. In this case, the maximal number of components is equal to the number of patterns in the dataset since each pattern can add at most one degree of freedom to the dataset $X \in \mathbb{R}^N$. Hence the maximum number of components $N_{\text{max}}$ of a pattern dataset $X$ is given as follows:
\begin{align}
    N_{\text{max}} = \min(N_{\text{values}}, N_{\text{patterns}}).
      \label{eq:nmax}
\end{align}
Here $N_{\text{values}} =512$ is the number of recorded intensity values per pattern and $N_{\text{patterns}}$ is the number of patterns in the dataset $X$. Fig~$\eqref{fig:components}$ below shows the cumulative explained variance ratio over the fraction of maximal No. components $N_{\text{max}}$ as defined above. In this figure, a faster convergence of the cumulative variance ratio towards one indicates that the patterns in this data are relatively similar. The degree of variation between the patterns is different for each contribution. For example, the CRNS and the HKUST contributions each are collections that encompass many research projects over a large period of time and thus exhibit a high degree of variability between individual patterns. In contrast, the contributions by USC and LBNL contain many very similar patterns. The patterns in the USC dataset are similar because the underlying samples are all variations of \ce{CuNi} and \ce{CuAl} alloys. The patterns submitted by LBNL are similar because they stem from in-situ recordings where several hundred or several thousand patterns were collected over time per sample while they were undergoing physical conversion processes.

@@

Fig.~$\eqref{fig:histograms}$ provides an overview of the distributions of pattern and structure properties in the opXRD database. Nearly all patterns have an angular resolution smaller than $\Delta(2\theta) = 0.1 ^\circ$. Here the angular resolution is defined as the range of recorded angles divided by the number of recorded intensity values along that range. For most patterns, the lowest recorded angle is smaller than $30 ^\circ$ and the highest recorded angle is smaller than $120 ^\circ$. The start-to-end angle distribution reveals that all diffractograms start in a narrow window between $0\degree$ and approximately $50\degree$, while they end between $50\degree$ and $150\degree$, with the majority of patterns going from $0\degree$ to approximately $70\degree$. Unlike most ML approaches using synthetic data over the full angle range with fixed resolution, the opXRD dataset has a strongly varying angle range and resolution. Hence, working with this data requires additional pre-processing methods such as padding and interpolation, or more flexible ML models beyond standard CNNs.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.8\linewidth]{figures/hist.png}
    \caption{Histograms detailing the distribution of pattern and structure properties in the opXRD database: a) distribution of spacegroups present in labeled data; b) distribution of angular resolution in all data; c) distribution of smallest and largest recorded $2\theta$ values for all data.}
    \label{fig:histograms}
\end{figure}

>>>>

\begin{table}[!htb]
\centering
\caption{\footnotesize Overview of the contributions to the opXRD database: The availability of the chemical composition, spacegroups, lattice parameters, and atomic coordinates of the underlying samples are indicated by the columns ``Comp.'', ``Spg.'', ``Lattice'' and ``\replaced{Basis}{ Atom coords}.'' respectively.}
\label{tab:merged}
\scalebox{0.82}{
\begin{adjustbox}{center}
\begin{NiceTabular}{@{}llllllll@{}}[columns-width=1.2cm]
%\begin{tabular}{@{}llccccl@{}}
\toprule
\textbf{\ Institution} & \textbf{No. patterns} & \textbf{Comp.} & \textbf{Spg.} & \textbf{Lattice} & \textbf{Basis} & \textbf{Research Project} \\
\midrule
\ CNRS        &  1052         &  \cmark   & \pchk{85}  & \cmark    &   \pchk{85}   & Diffraction data extracted from the COD \\ 
\ USC         &  338          &  \cmark   & \cmark     & \pchk{90} &   \xmark   & Study of CuNi and CuAl alloys          \\
\ HKUST(GZ)   &  520          &  \pchk{4} & \pchk{4}   & \pchk{4}  &   \pchk{4}   & Phase identification dataset  \\ 
\ EMPA        &  770          & \cmark    & \pchk{63}  & \xmark    &   \xmark   & Metal halide perovskites, Zn-V-N libraries \\ 
\ INT         &  19,796       & \xmark    & \xmark     & \xmark    &   \xmark   & Compilation of various projects        \\ 
\ IKFT        &  64           & \xmark    & \xmark     & \xmark    &   \xmark   & Commercial catalysts, metals, metal oxides   \\ 
\ LBNL        &  70,012       & \xmark    & \xmark     & \xmark    &   \xmark   & Perovskites precursors, Mn-Sb-O system  \\ 
\midrule
\midrule
\added{\ $\sum$ Labeled}   & 2,179  & \pchk{79}  & \pchk{66}  & \pchk{63}  & \pchk{42} & Partially and fully labeled opXRD data \\
\added{\ $\sum$ Unlabeled }& 90,373 & \xmark     & \xmark     & \xmark     & \xmark    & Unlabeled opXRD data\\
\bottomrule
\end{NiceTabular}
\end{adjustbox}}
\end{table}

The variance of the data was analyzed using principal component analysis (PCA). PCA can be applied to datasets $X \subset \mathbb{R}^N$ to reduce the number of components needed to describe points $p \in X$ up to some tolerance in lost accuracy. In the context of PCA, the cumulative explained variance ratio is a measure of how much of the variance in the dataset $X$ can be explained using a given number of components. For a rigorous definition of PCA and the explained variance ratio, we refer to the literature \cite{Jolliffe2016}. Here, PCA was performed on datasets of X-ray diffraction patterns. These datasets $X$ are subsets of $\mathbb{R}^{N}$ with $N=512$ since each pattern $p \in X$ was standardized to have 512 intensity values spread out evenly from $0\degree$ to $180\degree$ using zero padding and interpolation with cubic splines. Hence the maximal components that could be needed to describe a dataset of diffraction data in this context is $N=512$. However, the maximal number of components is even lower for datasets that contain less than 512 patterns. In this case, the maximal number of components is equal to the number of patterns in the dataset since each pattern can add at most one degree of freedom to the dataset $X \in \mathbb{R}^N$. Hence the maximum number of components $N_{\text{max}}$ of a pattern dataset $X$ is given as follows:
\begin{align}
    N_{\text{max}} = \min(N_{\text{values}}, N_{\text{patterns}}).
      \label{eq:nmax}
\end{align}
Here $N_{\text{values}} =512$ is the number of recorded intensity values per pattern and $N_{\text{patterns}}$ is the number of patterns in the dataset $X$. Fig~$\eqref{fig:components}$ below shows the cumulative explained variance ratio over the fraction of maximal No. components $N_{\text{max}}$ as defined above. In this figure, a faster convergence of the cumulative variance ratio towards one indicates that the patterns in this dataset are \replaced{largely}{relatively} similar. \deleted{The degree of variation between the patterns is different for each contribution.} \replaced{For example}{In contrast}, the contributions by USC and LBNL contain many very similar patterns. The patterns in the USC dataset are similar because the underlying samples are all variations of \ce{CuNi} and \ce{CuAl} alloys. The patterns submitted by LBNL are similar because they stem from in-situ recordings where several hundred or several thousand patterns were collected over time per sample. \deleted{ while they were undergoing physical conversion processes.} \replaced{In contrast}{For example}, the CRNS and the HKUST contributions each are collections that encompass many research projects over a large period of time and thus exhibit a high degree of variability between individual patterns.

@@

Fig.~$\eqref{fig:hist_XRD}$ \replaced{illustrates}{provides an overview of} the distributions of pattern \deleted{and structure} properties in the opXRD database. Nearly all patterns have an angular resolution smaller than $\Delta(2\theta) = 0.1 ^\circ$. Here the angular resolution is defined as the range of recorded angles divided by the number of recorded intensity values along that range. For most patterns, the lowest recorded angle is smaller than $30 ^\circ$ and the highest recorded angle is smaller than $120 ^\circ$. The start-to-end angle distribution reveals that all diffractograms start in a narrow window between $0\degree$ and approximately $50\degree$, while they end between $50\degree$ and $150\degree$, with the majority of patterns going from $0\degree$ to approximately $70\degree$. Unlike most ML approaches using synthetic data over the full angle range with fixed resolution, the opXRD dataset has a strongly varying angle range and resolution. Hence, working with this data requires additional pre-processing methods such as padding and interpolation, or more flexible ML models beyond standard CNNs.

\added{Fig.~$\eqref{fig:hist_CRYSTAL}$ illustrates how structural properties are distributed among the structures underlying the labeled subset of the opXRD database. Most structures include either \ce{N}, \ce{C}, or \ce{O} atoms, and have unit cells that contain less than 100 atoms and are smaller than $10^3$\AA. The most common spacegroups include the orthorhombic $\text{Pnma}$, the monoclinic $\text{P}2_1/c$, and the cubic $\text{Fm}3\text{m}$.}

\begin{figure}[!htb]
    \centering
    \includegraphics[width=\linewidth]{figures/hist_XRD.png}
    \caption{Histograms detailing properties of all diffraction patterns in the opXRD database: a) distribution of angular resolutions; b) distribution of smallest and largest recorded $2\theta$ values.}
    \label{fig:hist_XRD}
\end{figure}

\begin{figure}[!htb]
    \centering
    \includegraphics[width=\linewidth]{figures/hist_CRYSTAL.png}
    \caption{Histograms detailing properties of the structures underlying labeled diffraction patterns in the opXRD database: a) distribution of structures containing specified elements;b) distribution of structures containing spacegroups; c) distribution of No. atoms $N_{\text{atom}}$ contained in unit cell; d) distribution of unit cell volume $V_{\text{uc}}$}
    \label{fig:hist_CRYSTAL}
\end{figure}

\pagebreak
